---
title: "20440-Project"
author: "Christine Davis, Bri Ko"
date: "3/29/2021"
output: html_document
---

```{r setup, include=FALSE}
rm(list = ls()) #remove stored variables
knitr::opts_chunk$set(echo = TRUE) #set knitr options
```

#Load packages

```{r load, include=FALSE}
#For BiocManager commands, uncomment the first time you run this code to install
#packages directly using BiocManager. All other packages should be loaded using
#Tools -> Install Packages
library(readr)
library(pheatmap)
library(dendsort)
library(viridis)
library(ggrepel)
library(ggplot2)
library(dplyr)
library(limma)
library(BiocManager)
#BiocManager::install("biomaRt")
library(dnapath)
library(factoextra)
library(org.Hs.eg.db)
#BiocManager::install("org.Mm.eg.db")
#BiocManager::install("DOSE")
library(DOSE)
library(ggnewscale)
library(org.Mm.eg.db)
library(githubinstall)
#githubinstall("dnapath")
library(dnapath)
#BiocManager::install("reactome.db")
#BiocManager::install("topGO")
library(reactome.db)
library(SeqNet)
library(topGO)
library(tidyverse)
#BiocManager::install("ReactomePA")
library(ReactomePA)
library(VennDiagram)
#BiocManager::install("ComplexHeatmap")
library(ComplexHeatmap)
library(circlize)
set.seed(1234)  # random seed for reproducibility
```

#Download and clean data

```{r import data}
# Import the data and look at the first six rows- gives data in log2 cpm
data <- read.csv(file = '../Data/ABC_WT_DKO.csv')
head(data)

# Pull out gene names, counts (log2 cpm)
genes <- data[,1]
counts <- data[,7:10]
rownames(counts) <- genes
counts <- t(counts)
rev_log2 <- function(x) {( 2^x ) }  #(--> change back to regular cpm)
counts1 <- sapply(counts[1,],rev_log2)
counts2 <- sapply(counts[2,],rev_log2)
counts3 <- sapply(counts[3,],rev_log2)
counts4 <- sapply(counts[4,],rev_log2)
counts <- data.frame(rbind(counts1, counts2, counts3, counts4))
rownames(counts) <- colnames(data[,7:10])

#Z-score gene counts, create list of sample types
X <- scale(counts, center = TRUE, scale = TRUE)
ID <- c("WT","WT","DKO","DKO")

#Repeat for human data
hdata <- read.csv(file = '../Data/Human_SLE_Ctrl.csv')
head(hdata)

# Pull out gene names, counts (log2 cpm)
hgenes <- hdata[,1:2] 
hcounts <- hdata[,4:ncol(hdata)]
rownames(hcounts) <- hgenes[,1]
hcounts <- hcounts[-which(is.na(hgenes[,2])),]
hgenes <- hgenes[-which(is.na(hgenes[,2])),]
hcounts <- hcounts[-which(grepl("2-Mar",hgenes[,2])),]
hgenes <- hgenes[-which(grepl("2-Mar",hgenes[,2])),]
rownames(hcounts) <- hgenes[,2]
hcounts <- t(hcounts)

#Z-score gene counts, create list of sample types
hX <- scale(hcounts, center = TRUE, scale = TRUE)

#Name samples by disease state
df_diagnosis <- data.frame(sample = rownames(hX))
df_diagnosis$diagnosis <- rep(NA, length = nrow(df_diagnosis))
df_diagnosis$diagnosis[which(grepl("HC", df_diagnosis$sample))] <- "Control"
df_diagnosis$diagnosis[which(grepl("SLE", df_diagnosis$sample))] <- "SLE"
hID <- df_diagnosis$diagnosis
hX <- hX[order(rownames(hX)),]

```

#PCA for visualization of samples

```{r PCA}
#To run for mouse vs human, change X, ID (mouse) --> hX, hID (human)

#Create pca model from Z-scored gene counts
pca <- prcomp(X)

## Add sample info to PCs and plot
labeled_PCs <- as.data.frame(cbind(ID, pca$x))
plt <- ggplot(labeled_PCs, aes(x = PC1, y=PC2, color=ID)) + 
  geom_point(shape = 19, size = 10, alpha = 0.8) +
  labs(title = "PCA of Mouse WT vs DKO ABC Transcriptomes",
       x = "PC1", y = "PC2", color = "Disease State") +
  scale_color_manual(values=c("DKO" = 'red', "WT" = 'cornflowerblue')) +
  theme_classic() +
  theme(axis.text = element_blank(),
        axis.title = element_text(size = 24),
        axis.ticks = element_blank(), 
        plot.title = element_text(hjust=0.5, size = 24),
        legend.title = element_text(size = 18),
        legend.text = element_text(size = 18))

#Print PCA to pdf
pdf(paste("../Figures/PCA_mouse",".pdf", sep = ""), 
    width = 10, height = 10) 
print(plt)
dev.off()

```

# Differential Expression Analysis ("DE")

```{r Differential Expression}

#Create identification matrix designating each sample as DKO or WT
matrix_mouse <- model.matrix(~0+ID)
matrix_human <- model.matrix(~0+hID)
colnames(matrix_mouse) <- c("DKO","WT")
colnames(matrix_human) <- c("Control","SLE")

#Fit a linear model of gene counts by the design matrix, 
#so each group is compared. Transpose cpm counts to align with design matrix.
fit_human <- lmFit(t(hcounts), matrix_human)
fit_mouse <- lmFit(t(counts), matrix_mouse)

#makeContrasts takes expression specifying contrasts, returns contrast matrix
contrasts_mouse <- makeContrasts(DKO - WT, levels=matrix_mouse)
contrasts_human <- makeContrasts(SLE - Control, levels=matrix_human)
#Compute estimated coefficients from linear model fit, contrasts
fit2_mouse <- contrasts.fit(fit_mouse, contrasts_mouse)
fit2_human <- contrasts.fit(fit_human, contrasts_human)

#Takes microarray linear fit model, returns t-statistics, F-statistics, and 
#log-odds of differential expression using empirical Bayes' moderation
fit2_mouse <- eBayes(fit2_mouse)
fit2_human <- eBayes(fit2_human)

#Check how many genes are significantly over-expressed or under-expressed
#between conditions, and which are equally expressed (~10% of genes are 
#significantly diff. expressed)
table(decideTests(fit2_mouse)) #returns # of over (+1) or under (-1) expressed genes
table(decideTests(fit2_human))
#Get table of top-ranked genes, make gene names explicit variable
full_results_mouse <- topTable(fit2_mouse, number=Inf, adjust.method="BH")
full_results_mouse <- tibble::rownames_to_column(full_results_mouse,"ID")
full_results_human <- topTable(fit2_human, number=Inf, adjust.method="BH")
full_results_human <- tibble::rownames_to_column(full_results_human,"ID")

p_cutoff <- 0.05
fc_cutoff <- 1
topN <- 100 #top N genes for annotation

#Reformat full results to have additional variables of interest, filter for 
#set parameters, plot volcano plot of DE and significance
full_results_human <- full_results_human %>% 
  mutate(Significant = adj.P.Val < p_cutoff, abs(logFC) > fc_cutoff ) %>% 
  mutate(Rank = 1:n(), Label = ifelse(Rank < topN, ID,"")) %>%
  mutate(logpval = -log10(adj.P.Val))
full_results_mouse <- full_results_mouse %>% 
  mutate(Significant = adj.P.Val < p_cutoff, abs(logFC) > fc_cutoff ) %>% 
  mutate(Rank = 1:n(), Label = ifelse(Rank < topN, ID,"")) %>%
  mutate(logpval = -log10(adj.P.Val))

#Plot human or mouse volcano plots by switching title, full_results_(desired species)
plt <- ggplot(full_results_human, aes(x = logFC, y = logpval, col=Significant,
                                      label=Label)) + 
  geom_point() + 
  geom_text_repel(col="black", max.overlaps = 10) +
  ggtitle("Differential Expression Analysis- Human Control vs SLE B Cells") +#WT vs DKO ABCs") + #
  xlab("Log fold-change") + ylab("-log(p-value)") +
  xlim(-50,50) + #ylim(-5, -1) +
  theme_light() +
  theme(plot.title = element_text(hjust=0.5, size = 18))  

#Print volano to pdf
pdf(paste("../Figures/Volcano_Human",".pdf", sep = ""), 
    width = 10, height = 8) 
print(plt)
dev.off()

```
#Which genes are differentially expressed in both mouse and human models of SLE?
#What Reactome pathways are those genes enriched in?

```{r DE overlaps}

#Which pathways and genes are differentially expressed in both the mouse and human models?

#Make sure they are in the same direction! ie control > or < SLE in BOTH
#Both must also be significantly differentially expressed
higher_de_mouse <- subset(full_results_mouse, logFC > 0 & Significant == TRUE)
lower_de_mouse <- subset(full_results_mouse, logFC < 0 & Significant == TRUE)
higher_de_human <- subset(full_results_human, logFC > 0 & Significant == TRUE)
lower_de_human <- subset(full_results_human, logFC < 0 & Significant == TRUE)

high_de_overlap <- higher_de_human[which(toupper(higher_de_mouse$ID) %in% higher_de_human$ID),]
low_de_overlap <- lower_de_human[which(toupper(lower_de_mouse$ID) %in% lower_de_human$ID),]
de_overlap <- rbind(high_de_overlap, low_de_overlap)

#Use Entrez IDs (column 2) from overlap pathways to identify Reactome enriched pathways
human_gene_IDs <- mapIds(org.Hs.eg.db, de_overlap$ID, 'ENTREZID', 'SYMBOL')
de_overlap$EntrezID <- human_gene_IDs
de_overlap <- de_overlap[!is.na(de_overlap$EntrezID), ]
gene <- de_overlap$EntrezID
de_reactome <- enrichPathway(gene=gene, 
                             organism = "human",
                             pvalueCutoff = 0.05,
                             pAdjustMethod = "BH",
                             readable=FALSE)

#Dotplot visualization of Reactome network terms
p <- dotplot(de_reactome, showCategory=30)
#Print to pdf
pdf(paste("../Figures/Reactome_DotPlot",".pdf", sep = ""), 
    width = 20, height = 8) 
print(p)
dev.off()

## Alternative Reactome network visualization
edox <- setReadable(de_reactome, 'org.Hs.eg.db', 'ENTREZID')

p3 <- cnetplot(edox, showCategory = 30, circular = TRUE, colorEdge = TRUE)

#Print to pdf
pdf(paste("../Figures/Reactome_Network",".pdf", sep = ""), 
    width = 20, height = 8) 
print(p3)
dev.off()



```


#Differential connectivity

``` {r DC analysis}
#Convert gene names to Entrez IDs
mouse_gene_names <- mapIds(org.Mm.eg.db, colnames(counts), 'ENTREZID', 'SYMBOL')
human_gene_names <- mapIds(org.Hs.eg.db, colnames(hcounts), 'ENTREZID', 'SYMBOL')

#Add entrez IDs to dataframe
mouse_Entrez_genes <- data.frame(counts)
colnames(mouse_Entrez_genes) <- mouse_gene_names
human_Entrez_genes <- data.frame(hcounts)
colnames(human_Entrez_genes) <- human_gene_names

#Get reactome pathways represented within genes measured for both mouse & human
react_paths_human <- get_reactome_pathways(species = "Homo sapiens")
react_paths_mouse <- get_reactome_pathways(species = "Mus musculus")

# Run dnapath using gene expression and group information for human data
results_human <- dnapath(human_Entrez_genes, 
                   pathway_list = react_paths_human, 
                   groups = df_diagnosis$diagnosis,
                   network_inference = run_pcor)

results_mouse <- dnapath(mouse_Entrez_genes, 
                   pathway_list = react_paths_mouse, 
                   groups = ID,
                   network_inference = run_corr,
                   method = "pearson")

#results_human <- filter_pathways(results_human, alpha_pathway = 0.01)
#results_mouse <- filter_pathways(results_mouse,alpha_pathway = 0.333) #limited by replicate number
# Setting seed allows for reproducible work
set.seed(123)
results_mouse <- rename_genes(results_mouse, to = "symbol", species = "Mus musculus",
                        dir_save = tempdir())
results_human <- rename_genes(results_human, to = "symbol", species = "Homo Sapiens",
                        dir_save = tempdir())

plot(results_human[[100]])

# Extract the two estimated association networks for the first pathway
nw <- get_networks(results_human[[100]])
# Plot the networks using the SeqNet::plot_network function.

# Plot the two networks (in separate plots)
g <- SeqNet::plot_network(nw[[1]])
SeqNet::plot_network(nw[[1]], compare_graph = g)
# Plot of the differential network for pathway 1.
# Use compare_graph to maintain same layout
SeqNet::plot(results_human[[1]], compare_graph = g)

#Get dnapath_list format into dataframe for both mouse and human
pathway <- list()
genes <- list()
dc_score <- list()
pval <- list()
mean_expr1 <- list()
mean_expr2 <- list()

for (path in 1:length(results_human))
{
  tmp <- summary(results_human[[path]], by_gene = TRUE)
  pathway <- append(pathway, tmp$pathway)
  genes <- append(genes, tmp$genes)
  dc_score <- append(dc_score, tmp$dc_score)
  pval <- append(pval, tmp$p_value)
  mean_expr1 <- append(mean_expr1, tmp$mean_expr1)
  mean_expr2 <- append(mean_expr2, tmp$mean_expr2)
}

df_human <- data.frame(pathway = unlist(pathway),
                       genes = unlist(genes),
                       dc_score = unlist(dc_score),
                       pval = unlist(pval),
                       mean_expr1 = unlist(mean_expr1),
                       mean_expr2 = unlist(mean_expr2))

pathway <- list()
genes <- list()
dc_score <- list()
pval <- list()
mean_expr1 <- list()
mean_expr2 <- list()

for (path in 1:length(results_mouse))
{
  tmp <- summary(results_mouse[[path]], by_gene = TRUE)
  pathway <- append(pathway, tmp$pathway)
  genes <- append(toupper(genes), tmp$genes)
  dc_score <- append(dc_score, tmp$dc_score)
  pval <- append(pval, tmp$p_value)
  mean_expr1 <- append(mean_expr1, tmp$mean_expr1)
  mean_expr2 <- append(mean_expr2, tmp$mean_expr2)
}

df_mouse <- data.frame(pathway = unlist(pathway),
                       genes = unlist(genes),
                       dc_score = unlist(dc_score),
                       pval = unlist(pval),
                       mean_expr1 = unlist(mean_expr1),
                       mean_expr2 = unlist(mean_expr2))

#Which genes are differentially connected in both the mouse and human models?

df_human_overlap <- unique(df_human$genes[which(df_human$genes %in% df_mouse$genes)])
df_mouse_overlap <- unique(df_mouse$genes[which(df_mouse$genes %in% df_human$genes)])
df_overlap <- data.frame(human = sort(df_human_overlap), mouse = sort(df_mouse_overlap))

#What human pathways are those overlapping genes involved in?
human_overlap <- df_human[which(df_human$genes %in% df_overlap$human),]
human_overlap <- human_overlap[which(human_overlap$pval < 0.05),]
human_overlap <- as.data.frame(count(human_overlap,human_overlap$pathway))
colnames(human_overlap) <- c("Pathway","Count")
human_overlap$Pathway <- sapply(strsplit(human_overlap$Pathway,"See"), `[`, 1)
human_overlap$Pathway <- as.factor(human_overlap$Pathway)


#Plot top 20 pathways (by # of DE genes shared by mouse/human)
#Most of #s15-25 are signalling, so selected these for showing
ordered_overlap <- human_overlap[order(-human_overlap$Count),]
ordered_overlap_a <- ordered_overlap[which(ordered_overlap$Count >5),]
ordered_overlap_b <- ordered_overlap[which(grepl("signal|Caspase|DUB"
                                                 ,ordered_overlap$Pathway)),] #adding to make an even 30
ordered_overlap_b <- ordered_overlap_b[which(ordered_overlap_b$Count == 5),]
oo <- rbind(ordered_overlap_a,ordered_overlap_b)

#Plot of differentially connected networks ordered by # of DC genes in network
p2 <-  ggplot(oo, aes(x=reorder(Pathway, Count), y=Count)) +
    geom_bar(stat="identity", fill="#f68060", alpha=.6) +
    coord_flip() +
    labs(y="Number of Significantly Differentially Connected Genes in Pathway",
         x = "Reactome Pathway",
         title = "Top 30 Human Pathways Most Differentially Connected") +
    theme_classic() +
    theme(axis.text = element_text(size = 40, face = "bold"),
        axis.title = element_text(size = 50),
        axis.ticks = element_blank(), 
        plot.title = element_text(hjust=0.5, size = 50))
p2

#Print plot to pdf
pdf(paste("../Figures/Reactome_DC",".pdf", sep = ""),
    width = 45, height = 40) 
print(p2)
dev.off() 

#Now only look at Reactome pathways from differentially connected AND expressed
df_human_sort <- data.frame(df_human[order(-df_human$dc_score),])
df_human_sort <- df_human_sort[which(
  abs(log2(df_human_sort$mean_expr1/df_human_sort$mean_expr2)) > 1),]
df_human_sort <- df_human_sort[-which(
  df_human_sort$pval > 0.05),]
path_list <- count(df_human_sort,df_human_sort$pathway)
df_human_sort_rank <- path_list[order(-path_list$n),]
colnames(df_human_sort_rank) <-c("Pathway","Count")
df_human_sort_rank$Pathway <- sapply(strsplit(df_human_sort_rank$Pathway,"See"), `[`, 1)
df_human_sort_rank$Pathway <- as.factor(df_human_sort_rank$Pathway)
  
p3 <-  ggplot(subset(df_human_sort_rank, Count >1),
              aes(x=reorder(Pathway, Count), y=Count)) +
    geom_bar(stat="identity", fill="#f68060", alpha=.6) +
    coord_flip() +
    labs(y="Number of Significantly Differentially Connected   \n and Enriched Genes in Pathway",
         x = "Reactome Pathway",
         title = "Top Human Pathways Most Enriched \n in Differentially Connected and Expressed Genes") +
    theme_classic() +
    theme(axis.text = element_text(size = 40, face = "bold"),
        axis.title = element_text(size = 50),
        axis.ticks = element_blank(), 
        plot.title = element_text(hjust=0.5, size = 50))

#Print plot to pdf
pdf(paste("../Figures/Reactome_DC_DE",".pdf", sep = ""),
    width = 45, height = 40) 
print(p3)
dev.off()
```

#DC Volcano plots

```{r DC Volcano plots}
#to plot human or mouse plots, change ggplot input human_overlap <-> mouse_overlap, and title
p_cutoff <- 0.34
dc_cutoff <- 0.1
topN <- 20 #top N genes for annotation

#Reformat full results to have additional variables of interest, filter for 
#set parameters, plot volcano plot of DE and significance
human_overlap <- human_overlap %>% 
  mutate(Significant = pval < p_cutoff, abs(dc_score) > dc_cutoff ) %>% 
  mutate(Rank = 1:n(), Label = ifelse(Rank < topN, genes,"")) %>%
  mutate(logpval = -log10(pval))
mouse_overlap <- mouse_overlap %>% 
  mutate(Significant = pval < p_cutoff, abs(dc_score) > dc_cutoff ) %>% 
  mutate(Rank = 1:n(), Label = ifelse(Rank < topN, genes,"")) %>%
  mutate(logpval = -log10(pval))

#Plot human or mouse volcano plots by switching title, full_results_(desired species)
plt <- ggplot(mouse_overlap, aes(x = dc_score, y = logpval, col=Significant,
                                      label=Label)) + 
  geom_jitter(alpha = 0.35) + #change to geom_point for human
  geom_text_repel(col="black", max.overlaps = 10) +
  ggtitle("Differential Connectivity Analysis- WT vs DKO ABCs") + #Human Control vs SLE B Cells") +#
  xlab("Differential Connectivity Score") + ylab("-log(p-value)") +
  ylim(-.1,1) + #ylim(-5, -1) +
  theme_light() +
  theme(plot.title = element_text(hjust=0.5, size = 18))  

#Print volano to pdf
pdf(paste("../Figures/Volcano_DC_Mouse",".pdf", sep = ""), 
    width = 10, height = 8) 
print(plt)
dev.off()


```

# Venn Diagrams for both DE and DC analyses

```{r Venn Diagrams}

#Get all gene names that are differentially expressed and higher in SLE, DE
#and lower in SLE, and differentially connected. Do for mouse and human.
mouse_comparisons <- cbind(higher_de_mouse$ID, lower_de_mouse$ID,
                          df_mouse$genes)
human_comparisons<- cbind(higher_de_human$ID, lower_de_human$ID,
                         df_human$genes)
filenames <- c("../Figures/DE_higher_Venn.png",
                  "../Figures/DE_lower_Venn.png",
                  "../Figures/DC_Venn.png")

#Plot Venn Diagram of DE overlaps, looping through each of 3 subsets
#(DE and higher expression in SLE, DE and lower expression in SLE, DC)
for (i in 1:3)
{
  mouse <- unique(mouse_comparisons[,i])
  human <- unique(human_comparisons[,i])
  file <- filenames[i]
  
  venn.diagram(
  x = list(
    human, 
    toupper(mouse) #make mouse names uppercase for matching
    ),
  category.names = c("Human" , "Mouse"),
  filename = file,
  output = TRUE ,
          imagetype="png" ,
          height = 680 , 
          width = 680 , 
          resolution = 500,
          compression = "lzw",
          lwd = 1,
          scaled = TRUE,
          col=c("#440154ff", '#21908dff'), #choose colors for Venn
          fill = c(alpha("#440154ff",0.3), alpha('#21908dff',0.3)),
          cex = 0.5,
          fontfamily = "serif",
          ext.text = TRUE, #allow text outside diagram
          cat.cex = c(0.6, 0.6), #scale category label size
          cat.default.pos = "text",
          cat.pos = c(-27, 180), #degree location of label around circles
          cat.dist = c(0.055, -0.07), #distance of label from circle edge
          cat.fontfamily = "serif",
          cat.col = c("#440154ff", '#21908dff'),
          ext.line.lwd = 0.8, #width of line to label
          ext.line.lty = 3, #make line dotted (2 is solid)
          ext.dist = c(0.015,-0.2), #distance of line to label
          ext.length = 0.8 #length of line to label as % of 1
        )
}

```

#Use complex heatmap to overlay 4 heatmaps: 1) mouse DE genes, 2) human DE genes, 3) mouse DC genes (2 row, use mean expr 1 and mean expr 2 z-scored by column), 4) human DC genes (same as mouse)

```{r Heatmap}

X_tmp <- t(X[,which(toupper(colnames(X)) %in% colnames(hX))])
X_tmp <- X_tmp[1:2000,]
hX_tmp <- t(hX[,which(colnames(hX) %in% toupper(rownames(X_tmp)))])

#Cluster genes using hierarchical clustering with average Euclidean distance
order_cluster_cols <- hclust(dist(t(X_tmp)), method="average")

#Define sorting function to sort dendrogram/clustering distances by magnitude
sort_hclust <- function(...) as.hclust(dendsort(as.dendrogram(...)))

#Sort columns and rows according to sorting function
mat_cluster_cols <- sort_hclust(order_cluster_cols)
mat_cluster_rows <- sort_hclust(hclust(dist(X_tmp)))

#Define quantile_breaks function to take given matrix, split into 10 equal
#pieces, and return break locations. This will be used to color the heatmap
#with each color representing 10% of the data, to distinguish deciles
quantile_breaks <- function(xs, n = 10) {
  breaks <- quantile(xs, probs = seq(0, 1, length.out = n), na.rm = TRUE)
  breaks[!duplicated(breaks)]
}

#Use break function to create breaks in gene expression data 
mat_breaks <- quantile_breaks(X_tmp, n = 10)

#Label heatmap rows with sample type, columns with gene names
annotation_col = data.frame(strain = ID)
rownames(annotation_col) <- colnames(X_tmp)

#Provide colors for sample types
my_colors <- list(
  strain = c("WT" = '#8C02CC', "DKO" = '#5192C3'),
  disease = c("SLE" = 'deeppink3', "Control" = 'darkviolet'),
  Differential_Expression_Mouse = c("TRUE" = 'aquamarine2', "FALSE" = 'darkgrey'),
  Differential_Expression_Human = c("TRUE" = 'cyan1', "FALSE" = 'darkgrey'),
  Differential_Connectivity_Mouse = colorRamp2(c(-1, 0.5, 2), c("peachpuff", "pink2", "deeppink3")),
  Differential_Connectivity_Human = colorRamp2(c(-0.1, 0.5, 2), c("lightyellow", "skyblue", "slateblue4")))

#generate heatmap of log2 cpm count for differentially expressed genes,
#using sample color assignment, clustering columns and rows in previously sorted 
#order from sort_hclust, breaks from previously defined mat_breaks,
#and providing specific font,cell, and dedrogram (tree) sizes

p1 <- ComplexHeatmap::pheatmap(X_tmp, color = viridis(9), 
         annotation_colors = my_colors, 
         annotation_col = annotation_col, 
         cellwidth = 18, 
         cellheight = .2, 
         fontsize = 0, 
         border_color = NA, 
         treeheight_row = 3, 
         treeheight_col = 12,
         breaks = mat_breaks,
         cluster_cols = mat_cluster_cols,
         cluster_rows = mat_cluster_rows,
         )
#Print heatmap to pdf
pdf(paste("../Figures/Pheatmap_mouse",".pdf", sep = ""), 
    width = 20, height = 20) 
print(p1)
dev.off()

#Now build heatmap of human data that matches top selected mouse DEGs
hX_reorder <- hX_tmp[toupper(mat_cluster_rows$labels),]
annotation_col = data.frame(disease = sort(hID)) #make sure hID order correct
rownames(annotation_col) <- colnames(hX_reorder)
#uncomment below if you want to look at top human genes, not just overlap genes
#hX_reorder <- hX[,which(colnames(hX) %in% full_results_human$ID[1:2000])]

p2 <- ComplexHeatmap::pheatmap(hX_reorder, color = viridis(9), 
         annotation_colors = my_colors, 
         annotation_col = annotation_col, 
         cellwidth = 2, 
         cellheight = .2, 
         fontsize = 0, 
         treeheight_col = 0,
         border_color = NA, 
         cluster_cols = FALSE,
         breaks = mat_breaks,
         cluster_rows = FALSE
         )
#Print heatmap to pdf
pdf(paste("../Figures/Pheatmap_human",".pdf", sep = ""), 
    width = 20, height = 20) 
print(p2)
dev.off()

#Build heatmap of significance of expression differences in mice
Mouse_DEG_sig <- full_results_mouse[which(full_results_mouse$ID 
                                          %in% rownames(X_tmp)),]
rownames(Mouse_DEG_sig) <-Mouse_DEG_sig$ID
Mouse_DEG_sig_reorder <- Mouse_DEG_sig[mat_cluster_rows$labels,8]

#Build heatmap of significance of expression differences in humans
Human_DEG_sig <- full_results_human[which(full_results_human$ID 
                                          %in% toupper(rownames(X_tmp))),]
rownames(Human_DEG_sig) <-Human_DEG_sig$ID
Human_DEG_sig_reorder <- Human_DEG_sig[toupper(mat_cluster_rows$labels),8]

#Build heatmap of differential connectivity in mice
#df_mouse$dc_score, also include p.value
#df_human$dc_score, also include p.value

Mouse_DC_reorder <- df_mouse[which(df_mouse$genes %in% 
                                              toupper(mat_cluster_rows$labels)),]
Mouse_DC_reorder <- Mouse_DC_reorder[order(-Mouse_DC_reorder$dc_score),]
Mouse_DC_reorder[which(Mouse_DC_reorder$pval > 0.35),3] <- 0

dc_score_list_mouse <- list()
i = 0
for (gene in toupper(mat_cluster_rows$labels))
{
  i = i+1
  if (gene %in% Mouse_DC_reorder$genes)
  {
    tmp <- Mouse_DC_reorder[which(grepl(gene, Mouse_DC_reorder$genes)),3]
    dc_score_list_mouse[i] <- tmp[which.max(tmp)]
  } else
  {
    dc_score_list_mouse[i] <- 0
  }
  names(dc_score_list_mouse)[i] <- gene
}
dc_score_list_mouse <- unlist(dc_score_list_mouse)


#Repeat for human data
Human_DC_reorder <- df_human[which(df_human$genes %in% 
                                              toupper(mat_cluster_rows$labels)),]
Human_DC_reorder <- Human_DC_reorder[order(-Human_DC_reorder$dc_score),]
Human_DC_reorder[which(Human_DC_reorder$pval > 0.05),3] <- 0

dc_score_list_human <- list()
i = 0
for (gene in toupper(mat_cluster_rows$labels))
{
  i = i+1
  gene_list[i] <- gene
  if (gene %in% Human_DC_reorder$genes)
  {
    tmp <- Human_DC_reorder[which(grepl(gene, Human_DC_reorder$genes)),3]
    dc_score_list_human[i] <- tmp[which.max(tmp)]
  } else
  {
    dc_score_list_human[i] <- 0
  }
  names(dc_score_list_human)[i] <- gene
}
dc_score_list_human <- unlist(dc_score_list_human)

#Put all pieces together into complex heatmap

plt <- p1+p2 + rowAnnotation(Differential_Expression_Mouse = Mouse_DEG_sig_reorder,
                        col = my_colors) +
  rowAnnotation(Differential_Expression_Human = Human_DEG_sig_reorder,
                        col = my_colors) +
  rowAnnotation(Differential_Connectivity_Mouse = scale(dc_score_list_mouse, center = TRUE, scale = TRUE),
                col = my_colors) +
  rowAnnotation(Differential_Connectivity_Human = scale(dc_score_list_human, center = TRUE, scale = TRUE),
                col = my_colors)

#Print heatmap to pdf
pdf(paste("../Figures/Complex_Heatmap",".pdf", sep = ""), 
    width = 10, height = 15) 
print(plt)
dev.off()
```

```{r Complex Heatmap}
cite_packages()
```

